Anthropogenic plant communities of the Iberian Atlantic ecoregion

### 2024-04-10 doc/manuscript.Rmd
revised figures

Anthropogenic vegetation is the set of plant communities that occur as a direct consequence of human activities, and is chiefly composed by the weed vegetation of arable fields and the ruderal vegetation of human settlements and their surroundings [@RN5441]. The formation of anthropogenic vegetation in temperate Europe has been characterized by a constant enrichment produced by the arrival of alien species: first archaeophytes starting in the Neolithic and peaking in the Bronze Age; and then neophytes from the Modern Age to the present day [@RN5439]. Synanthropic archaeophytes include many Mediterranean taxa that started their expansion into temperate areas during the Mid-Holocene warm period, but it is difficult to disentangle to what extent this expansion was favored by human activities and/or ongoing climatic changes [@RN5432; @RN5443], especially in southern European areas close to the transition between the Mediterranean and the temperate climates [@RN5444]. Meanwhile, synanthropic neophytes include many American, African and Asian taxa that arrived as a consequence of the global trade networks established by the Western European colonial powers [@RN5451]. In the last century, anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer, and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

The anthropogenic communities of human-made habitats make up an important part of the supply of natural elements in urban and peri-urban habitats [@RN5445]. In densely-populated city landscapes, the plant communities that colonize urban vacant lots [@RN2300] are a useful biodiversity resource with high ecological and societal potential [@RN5446]. Anthropogenic vegetation cover can also impact regional-level extreme climatic events [@RN5440]. Furthermore, many synanthropic plants can accumulate heavy metals in their biomass [@RN5433], making them valuable candidates for nature-based solutions such as bioremediation and restoration of degraded landscapes [@RN5447].

The Iberian Atlantic ecoregion has had a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. This, together with its transitional position at the border between the temperate and Mediterranean climatic zones of Europe [@RN4700], suggests a long regional history for archaeophyte-rich anthropogenic plant communities, and supports a current high diversity of anthropogenic plant communities combining temperate and Mediterranean floristic elements [@RN5008]. Historical trade links with America and Asia, together with the warm temperate climate, have made the region a hotspot for plant invasions by neophytes [@Checklist]. Recent processes of postindustrial land-use change have left large areas of brownfields whose management and restoration requires an understanding of anthropogenic communities able to colonize or restore them [@RN5448; @RN5449]. Although there is a long tradition of studying weed and ruderal communities in the ecoregion [@RN5364; @RN5361; @RN5359; @RN5360] there is a current need for a regional-level synthesis that revises the classification and updates it in accord with recent developments in European vegetation science [@RN4698] and habitat classification [@RN5061]. In this article we have performed such a synthesis with two main objectives: (1) provide an updated classification of anthropogenic plant communities in the Iberian Atlantic ecoregion, and (2) describe the main patterns of variation in terms of species traits and ecological and disturbance preferences.

@@

We performed all data analyses with R version 4.3.1 [@RN5387], using the R package 'tidyverse' [@RN4662] for data processing and visualization. The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/manmade.

@@

As vegetation data for our analyses, we obtained all vegetation relevés (i.e. records of plant species coexisting in a given sampling plot at a given date) from our study ecoregion that were available in SIVIM, the Iberian and Macaronesian Vegetation Information System [@RN5374]. These were 28,775 relevés representing all vegetation types (i.e. anthropogenic, semi-natural and natural). Since these relevés were sampled by different authors throughout the decades, there was no taxonomical consistency among them, so we homogenized all taxon names for the whole dataset. We followed the methodology applied by Jiménez-Alfaro *et al.* [-@RN5375] to the Cantabrian Mountains, an area within the study ecoregion which includes c. 80% of the ecoregional flora, to create a new taxonomical checklist for the ecoregion based on the Euro+Med Plantbase [@RN5044]. For vascular plants not included in Euro+Med (e.g. several *Quercus* or *Salix* hybrids, non-native species, etc.), as well as for bryophytes, we followed Plants of the World Online [@RN5376]. We removed Fungi and algae. For analytical purposes, we merged taxa into aggregates (1) when there were sets of very close species with a difficult identification in the field (e.g. *Viola riviniana* + *V. reichenbachiana* + *V. sylvestris* = *Viola riviniana* aggr.), (2) when the taxon was recently separated in different taxa (e.g. *Hedera helix* + *Hedera hibernica* + *Hedera canariensis* = *Hedera helix* aggr.) and (3) when it was assumed that not all authors were able to identify or recognize infraspecific taxa (e.g. all *Narcissus pseudonarcissus* subspecies and varieties = *Narcissus pseudonarcissus* aggr.).

For mapping purposes, we improved the geographical information of the SIVIM relevés using a downscaling procedure. The original position of vegetation relevés in SIVIM was recorded as 10 km x 10 km (52.2%) and 1 km x 1 km (47.8%) grid cells. We assigned more precise geographic coordinates to these plots using the elevation recorded originally by the authors and a digital elevation model (DEM) of the Iberian Peninsula at 200 m grid resolution. From the original 10 km x 10 km or 1 km x 1 km grid cell of each relevé, we (1) kept only those 200 m cells that agreed with the relevé's elevation (± 50 m); (2) randomly selected one of these 200 m cells; and (3) assigned to the relevé the coordinate of the 200 m cell's centroid, in decimal degrees. By doing this, the plots were downscaled at higher resolution within their cells, assuming that 200 m cells at similar elevations within the same grid cell also have similar climatic conditions.

To identify anthropogenic relevés in the SIVIM database we followed two complementary approaches. Fist, we used the expert system created to classify vegetation plots into EUNIS pan-European habitat classification types [@RN5061]. The expert system uses the floristic composition of plots to assign them to habitat types depending on the presence and abundance of characteristic species. We used the habitats assigned by the expert system to retrieve plots that had been assigned to any habitat related to the vegetation classes we had previously defined: all habitats in the level 1 code V (vegetated man-made habitats), plus level 3 codes R55 (lowland moist or wet tall-herb and fern fringe) and R57 (herbaceous forest clearing vegetation). Second, we also retrieved relevés if they had been classified by the original authors of the relevé into any sintaxa belonging to our target anthropogenic vegetation classes (**Supplementary 1**). This produced an initial pool of 3,160 relevés from SIVIM, to which we added 89 relevés of urban or peri-urban plant communities conducted by us or extracted by us from the local literature [@RN5377; @RN5378]. After initial exploration of this dataset using Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379], we identified 652 outlier relevés which either did not belong to the target anthropogenic classes or to the geographical area of the ecoregion. We removed these, leaving a dataset of 2,508 relevés for the analyses of this article: 2,419 relevés from SIVIM, originally recorded in 89 publications see (**Supplementary 2**); plus 89 relevés added by us.

@@

[Definition of target alliances and training dataset.]{.underline} As a previous step before classification, we prepared a revised checklist of potential anthropogenic alliances that could be present in our study ecoregion, based on regional sintaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. Then, we filtered our dataset to include only those relevés which had been assigned by the authors to any of these alliances, and used sequential TWINSPANs to do a preliminary re-classification of these filtered relevés, attempting to re-match the relevés into the *a priori* defined checklist of alliances. We used 3 pseudospecies cut levels (0, 15, 25) with a minimum group size of 10 plots and Sørensen's average dissimilarity, and the R package 'twinspanR' [@RN5380] to do the TWINSPANs. To start, we did a TWINSPAN classification into 15 clusters, followed by separate TWINSPANS within each cluster. This preliminary TWINSPAN allowed us to (1) validate which alliances were present in the study ecoregion; and (2) prepare a training dataset for semi-supervised classification (see below).

[Semi-supervised classification.]{.underline} Next, we applied a semi-supervised classification to the whole dataset. Semi-supervised classification uses a training subset of *a priori* classified relevés to classify a secondary subset of unclassified relevés [@RN5383]. Our training subset included all the relevés that (1) had been classified by the authors of the relevé into any of the alliances that we had validated during the previous TWINSPANs, and (2) had also been classified by us into these alliances during our TWINSPANs. The secondary subset included all the relevés which did not meet these two criteria. Our goal was to refine and not to simply extend the classification, so we applied the semi-supervised classification to the whole dataset, i.e. we allowed relevés from the training subset to be re-assigned to other groups during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already *a priori* groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows the relevés to be classified either into the *a priori* groups or into a 'noise' group which includes outliers and transitional relevés [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional relevés; and we set the distance to the noise class to d = 1. We performed this analysis with the R package 'vegclust' [@RN5383]. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package 'RVAideMemoire' [@RN5381] to test the significant of the final vegetation groups; and Detrended Correspondence Analysis (DCA) as implemented in the R package 'vegan' [@RN5386] to visualize the relationships between the groups.

[Characteristic species and EUNIS habitat regionalization.]{.underline} We calculated sets of characteristic species [@RN5061] for each of the final vegetation groups. We defined dominant species as species with more than 25% cover in at least 5% of the relevés of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species with an 'IndVal' value with a p-value lower than 0.05, as calculated with the R package 'labdsv' [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each group a regionalized level 4 EUNIS habitat code, using pre-existing codes when they existed, and creating new codes when not.

[Community traits and ecological requirements.]{.underline} To describe the resulting vegetation groups, we compiled a dataset of species traits and ecological preferences for the species pool of the classified relevés. For each species, we extracted from *Flora iberica* [-@RN3243] its maximum height, its median month of flowering, the length of the flowering period, and whether it was a neophyte or not. From FloraVeg.EU [-@RN5385] we extracted the species’ life forms and kept 'therophyte' and 'geophyte' for further analysis since these two categories had the largest contribution to variance as per an exploratory Principal Component Analysis (PCA) performed using the R package 'FactoMineR' [@RN3166]. We also extracted the ecological indicator values [@RN5384] and the disturbance indicator values [@RN5101] of each species. In the case of the ecological and disturbance indicators, we assigned values to species lacking them by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, for each relevé we calculated its community-weighted mean (weighting by species cover) for each trait and ecological preference, and explored the main axes of variation using PCA, separately for species traits and ecological preferences.

@@

Our revision of the literature produced a provisional list of 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion (**Supplementary 3**). In our study dataset (n = 2,508 relevés), there were 2,201 relevés that had been assigned by the authors to any of these alliances. We used these 2,201 relevés for a preliminary re-classification based on TWINSPAN, attempting to re-match the relevés into the *a priori* defined checklist of 38 alliances. This procedure allowed us to validate 28 alliances as present in the ecoregion. A total of 1,725 relevés had been classified by both the original authors and our TWINSPAN into these alliances, and these became our training subset for semi-supervised classification. Attempts by the semi-supervised classification to create new groups resulted on new groups with no ecological significance, and thus we kept the 28 alliances as final vegetation groups (**Table 1**). Most of the alliances had a wide distribution in the ecoregion according to our dataset (**Figure 1**), although five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis*, *Paspalo-Agrostion semiverticillati*) had only isolated occurrences in a few points towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Supplementary 4**) within these alliances. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 relevés into 28 alliances and 427 relevés left in the noise group.

Focusing on the classified plots, the dataset included 1,167 taxa or taxa aggregates. Of these, 101 were neophytes (i.e. 9% of the species pool). Regarding life forms, the species pool was dominated by hemicryptophytes (43%) and therophytes (41%). The 10 most frequent species were *Ochlopoa annua* (621 occurrences), *Urtica dioica* (592), *Sonchus oleraceus* (540), *Stellaria media* (516), *Capsella bursa-pastoris* (390), *Polygonum aviculare* (375), *Dactylis glomerata* (363), *Senecio vulgaris* (332) and *Anisantha sterilis* (301). Considering only those relevés with the most frequent plot size (20 m^2^, n = 256 relevés), the average species richness per plot was 16 (minimum = 3, maximum = 37).

DCA ordination of the floristic composition of the classified relevés (**Fig. 2**) suggested a high consistency of the vegetation classes *Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea* and *Bidentetea*. The alliances belonging to the two perennial ruderal vegetation classes (*Artemisietea vulgaris* and *Epilobietea angustifolii*) were mixed together. The winter-annual class *Chenopodietea* also showed high heterogeneity, specially because of the isolation of the altered dune alliance *Linario polygalifoliae-Vulpion alopecuri*.

@@

PCA ordination of the community-weighted means of the species ecological and disturbance preferences (**Fig. 3B**) indicated that the first two axes explained 56% of the variability. The major contributors to axis 1 (35% variance explained) were disturbance frequency, soil disturbance, mowing frequency and light requirements; it separated communities thriving with frequent disturbances in open situations (classes *Polygono-Poetea annuae*, *Sisymbrietea*, *Digitario-Eragrostietea* and *Papaveretea rhoeadis*) from communities preferring less frequent disturbances and more shade (classes *Cymbalario-Parietarietea diffusae*, *Epilobietea angustifolii* and *Bidentetea*). The major contributors to axis 2 (21% variance explained) were nutrient requirements and disturbance severity; it separated more nitrophilous communities adapted to more severe disturbances (classes *Digitario-Eragrostietea*, *Sisymbrietea* and *Epilobietea angustifolii*) from less nitrophilous communities adapted to lower intensity disturbances (classes *Cymbalario-Parietarietea diffusae*, *Artemisietea vulgaris* and *Chenopodietea*).

@@

@@

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Figure 1. Distribution of the anthropogenic alliances in the Iberian Atlantic ecoregion. Each dot represents a relevé. Dots are colored by vegetation classes."}
knitr::include_graphics("../results/figures/maps.png")
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Figure 2. Ordination of the floristic composition of the anthropogenic plant communities of the Iberian Atlantic ecoregion. The ordination is the results of a Detrended Correspondence Analysis grouped by vegetation classes (A, left) and alliances (B, right)."}
knitr::include_graphics("../results/figures/dca.png")
```

```{r fig3, echo = FALSE, fig.pos = "H", fig.cap = "Figure 3. Ordination of the community-level species traits and ecological preferences. The ordinations are the results of a Principal Component Analysis calculated for the community-weighted means for species traits (A, left) and ecological and disturbance indicator values (B, right)."}
knitr::include_graphics("../results/figures/cwms.png")
```

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/boxplots.png")
```

>>>>

Anthropogenic vegetation is the set of plant communities that occur as a direct consequence of human activities, and is chiefly composed by the weed vegetation of arable fields and the ruderal vegetation of human settlements and their surroundings [@RN5441]. The anthropogenic communities of human-made habitats make up an important part of the supply of natural elements in urban and peri-urban habitats [@RN5445]. In densely-populated city landscapes, the plant communities that colonize urban vacant lots [@RN2300] are a useful biodiversity resource with high ecological and societal potential [@RN5446]. Anthropogenic vegetation cover can also impact regional-level extreme climatic events [@RN5440]. Furthermore, many synanthropic plants can accumulate heavy metals in their biomass [@RN5433], making them valuable candidates for nature-based solutions such as bioremediation and restoration of degraded landscapes [@RN5447].

The anthropogenic vegetation of temperate Europe is formed by an original pool of native species favored by humans, but this pool has been constantly enriched by the arrival of alien species: first archaeophytes starting in the Neolithic and peaking in the Bronze Age; and then neophytes from the Modern Age (after 1.500 A.D.) to the present day [@RN5439]. Synanthropic archaeophytes include many Mediterranean taxa that started their expansion into temperate areas during the Mid-Holocene warm period, but it is difficult to disentangle to what extent this expansion was favored by human activities and/or ongoing climatic changes [@RN5432; @RN5443], especially in southern European areas close to the transition between the Mediterranean and the temperate climates [@RN5444]. Meanwhile, synanthropic neophytes include many American, African and Asian taxa that arrived as a consequence of the global trade networks established by the Western European colonial powers [@RN5451]. In the last century, anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer, and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

Antropogenic vegetation is likely to be similar within ecological regions (ecoregions, hereafter), since they are assumed to encompass areas with a similar biogeographic history and recurrent local ecosystems [@RN5491]. However, ecoregional-scale studies on the diversity of antropogenic vegetation are scarce in the literature. The Iberian Atlantic ecoregion (a.k.a. Cantabrian Mixed Forests ecoregion; NW Iberian Peninsula) has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. This, together with its transitional position at the border between the temperate and Mediterranean climatic zones of Europe [@RN4700], suggests a long regional history for archaeophyte-rich anthropogenic plant communities, likely supporting a current high diversity of anthropogenic plant communities with both temperate and Mediterranean floristic elements [@RN5008]. Historical trade links with America and Asia, together with the warm and humid temperate climate, have made the region a hotspot for neophyte invasions [@RN5492; @RN5493]. Recent post-industrial land-use changes have left large areas with abandoned industrial sites (i.e. brownfields) whose management and restoration requires an understanding of anthropogenic communities able to colonize or restore them [@RN5448; @RN5449]. Although there is a long tradition of studying weed and ruderal communities in the ecoregion [@RN5364; @RN5361; @RN5359; @RN5360] there is a current need for an ecoregional-level synthesis that revises and updates the classification and updates in accord with recent developments in European vegetation science [@RN4698] and habitat classification [@RN5061]. In this article we have performed such a synthesis with two main objectives: (1) provide an updated classification of anthropogenic plant communities in the Iberian Atlantic ecoregion, and (2) describe the main patterns of vegetation diversity in terms of floristic origin (natives, archaeophytes, neophytes); life forms; functional traits (plant height and flowering time); and ecological indicator values (temperature, moisture, light, nutrients, soil reaction, disturbance frequency and disturbance severity).

@@

@@

We retrieved anthropogenic vegetation plots from SIVIM, the Iberian and Macaronesian Vegetation Information System [@RN5374]. To identify plots as anthropogenic we followed two criteria: (1) We used the expert system created by Chytrý *et al.* [-@RN5061] to classify vegetation plots into EUNIS pan-European habitat types. This expert system uses the floristic composition of plots to assign them to habitat types depending on the presence and abundance of characteristic species. We used the habitats assigned by the expert system to retrieve plots that had been assigned to any habitat related to the vegetation classes we had previously defined: all habitats in the level 1 code V (vegetated man-made habitats), plus level 3 codes R55 (lowland moist or wet tall-herb and fern fringe) and R57 (herbaceous forest clearing vegetation) (R55 and R57 include communities that are classified as "grasslands" by EUNIS but as "anthropogenic" by by Mucina *et al.* [-@RN4698]). (2) We also retrieved vegetation plots from SIVIM if they had been classified by the original authors of the relevé into any sintaxa belonging to our target anthropogenic vegetation classes (**Supplementary 1**). This produced an initial pool of 3,160 vegetation plots from SIVIM, to which we added 89 vegetation plots of urban or peri-urban plant communities sampled by us or extracted from local literature not in SIVIM [@RN5377; @RN5378]. After initial exploration of this dataset using Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379], we identified 652 outlier plots which either did not belong to the target anthropogenic classes or to the geographical area of the ecoregion. The majority of these plots corresponded to coastal plant communities and had been misclassified by the expert system. We removed these, leaving a dataset of 2,508 vegetation  plots for data analyses: 2,419 plots from SIVIM, originally recorded in 89 publications (**Supplementary 2**); plus 89 plots added by us.

@@

We performed all data analyses with R version 4.3.1 [@RN5387], using the R package 'tidyverse' [@RN4662] for data processing and visualization. The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/manmade. We homogenized all taxon names in the analyses and the article using Euro+Med [@RN5044] or Plants of the World Online for taxa not in Euro+Med [@RN5376].

[Definition of target alliances and training dataset.]{.underline} As a previous step before classification, we prepared a revised checklist of potential anthropogenic alliances that could be present in our study ecoregion, based on regional sintaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. Then, we filtered our dataset to extract a subset with the vegetation plots which had been assigned by the authors to any of these alliances. We used TWINSPAN to conduct a preliminary numerical classification of this subset, attempting to match the plots into the *a priori* defined checklist of alliances. We used 3 pseudospecies cut levels (0, 15, 25) with a minimum group size of 10 plots and Sørensen's average dissimilarity, and the R package 'twinspanR' [@RN5380] to do the TWINSPANs. To start, we did a TWINSPAN classification into 15 clusters, followed by separate TWINSPANS within each cluster. This preliminary TWINSPAN allowed us to (1) validate which alliances were present in the study ecoregion; and (2) prepare a training dataset for semi-supervised classification (see below).

[Semi-supervised classification.]{.underline} Next, we applied a semi-supervised classification to the whole dataset. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Our training subset included all the plots that (1) had been classified by the authors of the relevé into any of the alliances that we had validated during the previous TWINSPANs, and (2) had also been matched by us into these alliances during our TWINSPANs. The secondary subset included all the plots which did not meet these two criteria. Our goal was to refine and not to simply extend the classification, so we applied the semi-supervised classification to the whole dataset, i.e. we allowed plots from the training subset to be re-assigned to other groups during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows the plots to be classified either into the *a priori* groups or into a 'noise' group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package 'vegclust' [@RN5383]. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package 'RVAideMemoire' [@RN5381] to test the significance of the final vegetation groups; along with Detrended Correspondence Analysis (DCA) as implemented in the R package 'vegan' [@RN5386] to visualize the relationships between the groups.

[Characteristic species and EUNIS habitat regionalization.]{.underline} We calculated sets of characteristic species [@RN5061] for each of the final vegetation groups. We defined dominant species as species with more than 25% cover in at least 5% of the vegetation plots of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species with an 'IndVal' value with a p-value lower than 0.05, as calculated with the R package 'labdsv' [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each group a regionalized level 4 EUNIS habitat code, using pre-existing codes when they existed (https://eunis.eea.europa.eu/habitats-code-browser-revised.jsp), and creating new codes when not.

[Community traits and ecological requirements.]{.underline} To describe the resulting vegetation groups, we compiled a dataset of species traits and ecological preferences for the species pool of the classified vegetation plots. For each species, we extracted from *Flora iberica* [-@RN3243] its maximum height, its median month of flowering, the length of the flowering period, and whether it was a neophyte or not. From FloraVeg.EU [-@RN5385] we extracted the species’ life forms and kept 'therophyte' and 'geophyte' for further analysis since these two categories had the largest contribution to variance as per an exploratory Principal Component Analysis (PCA) performed using the R package 'FactoMineR' [@RN3166]. We also collected the ecological indicator values [@RN5384] and the disturbance indicator values [@RN5101] of each species. In the case of the ecological and disturbance indicators, we assigned values to species lacking them by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, we calculated the plot-level community-weighted mean (weighting by species cover) for each trait and ecological preference, and explored the main axes of variation using PCA, separately for species traits and ecological preferences.

@@

Our revision of the literature produced a provisional list of 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion (**Supplementary 3**). In our study dataset (n = 2,508 vegetation plots), there were 2,201 plots that had been assigned by the original authors to one of these alliances. We used these 2,201 plots for a preliminary re-classification based on TWINSPAN, attempting to re-match the plots into the *a priori* defined checklist of 38 alliances. This procedure allowed us to validate 28 alliances as present in the ecoregion. A total of 1,725 plots had been classified by both the original authors and our TWINSPAN into these alliances, and these became our training subset for the semi-supervised classification. Attempts by the semi-supervised classification to create new groups resulted in new groups with no ecological significance, and thus we kept the original 28 alliances as final vegetation groups (**Table 1**). Most of the alliances had a wide distribution in the ecoregion according to our dataset (**Figure 1**), although five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis*, *Paspalo-Agrostion semiverticillati*) had only isolated occurrences in a few points towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Supplementary 4**) within these alliances. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group.

Focusing on the classified plots, the dataset included 1,167 taxa or taxa aggregates. Of these, 101 were neophytes (i.e. 9% of the species pool). Regarding life forms, the species pool was dominated by hemicryptophytes (43%) and therophytes (41%). The 10 most frequent species were *Ochlopoa annua* (621 occurrences), *Urtica dioica* (592), *Sonchus oleraceus* (540), *Stellaria media* (516), *Capsella bursa-pastoris* (390), *Polygonum aviculare* (375), *Dactylis glomerata* (363), *Senecio vulgaris* (332) and *Anisantha sterilis* (301). Considering only those plots with the most frequent plot size (20 m^2^, n = 256 plots), the average species richness per plot was 16 (minimum = 3, maximum = 37).

DCA ordination of the floristic composition of the classified plots (**Fig. 2**) suggested a high consistency of the vegetation classes *Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea* and *Bidentetea*. The alliances belonging to the two perennial ruderal vegetation classes (*Artemisietea vulgaris* and *Epilobietea angustifolii*) were mixed together. The winter-annual class *Chenopodietea* also showed high heterogeneity, especially because of the isolation of the altered dune alliance *Linario polygalifoliae-Vulpion alopecuri*.

@@

PCA ordination of the community-weighted means of the species ecological and disturbance preferences (**Fig. 3B**) indicated that the first two axes explained 56% of the variability. The major contributors to axis 1 (35% variance explained) were disturbance frequency, soil disturbance, mowing frequency and light requirements; it separated communities thriving with frequent disturbances in open situations (classes *Polygono-Poetea annuae*, *Sisymbrietea*, *Digitario-Eragrostietea* and *Papaveretea rhoeadis*) from communities preferring less frequent disturbances and shadier areas (classes *Cymbalario-Parietarietea diffusae*, *Epilobietea angustifolii* and *Bidentetea*). The major contributors to axis 2 (21% variance explained) were nutrient requirements and disturbance severity; it separated more nitrophilous communities adapted to more severe disturbances (classes *Digitario-Eragrostietea*, *Sisymbrietea* and *Epilobietea angustifolii*) from less nitrophilous communities adapted to lower intensity disturbances (classes *Cymbalario-Parietarietea diffusae*, *Artemisietea vulgaris* and *Chenopodietea*).

@@

Segetal vegetation of Palestine [@RN5494].

@@

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Figure 1. Distribution of the anthropogenic alliances in the Iberian Atlantic ecoregion. Each dot represents a relevé. Dots are colored by vegetation classes."}
knitr::include_graphics("../results/figures/F1 - maps.png")
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Figure 2. Ordination of the floristic composition of the anthropogenic plant communities of the Iberian Atlantic ecoregion. The ordination is the results of a Detrended Correspondence Analysis grouped by vegetation classes (A, left) and alliances (B, right)."}
knitr::include_graphics("../results/figures/F2 - pca.png")
```

```{r fig3, echo = FALSE, fig.pos = "H", fig.cap = "Figure 3. Ordination of the community-level species traits and ecological preferences. The ordinations are the results of a Principal Component Analysis calculated for the community-weighted means for species traits (A, left) and ecological and disturbance indicator values (B, right)."}
knitr::include_graphics("../results/figures/F3 - origins.png")
```

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F4 - lifeforms.png")
```

```{r fig5, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F5 - height-flowering.png")
```

```{r fig6, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F6 - indicators.png")
```

```{r fig7, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F7 - pca.png")
```
